Degeneracy in a restricted Boltzmann machine

### 2016-02-20 presentations/coda2016/coda2016_poster.tex
left side complete

%%tikz setup
\usepackage{tikz, subfig, amsthm}
\usepackage{tikz-3dplot}
\usetikzlibrary{arrows, shapes, positioning, backgrounds}

%%other options
\definecolor{isublue}{RGB}{58,117,196}
\definecolor{isured}{RGB}{206,17,38}
\counterwithout{figure}{chapter}
\counterwithout{table}{chapter}
\usepackage{amsmath}
\setlength{\parindent}{0cm}
\def\vdotswide{\vbox{\baselineskip=20pt \lineskiplimit=0pt \kern6pt \hbox{.}\hbox{.}\hbox{.}}}

@@

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\input{images/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), which consists of two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer. Hidden nodes indicated by gray circles and the visible nodes indicated by white circles \cite{friedman2001elements}.}
  \label{fig:rbm}
\end{figure}

\begin{figure}[ht]
  \centering
  \input{images/visibles.tikz}
  \caption{Visibles diagram.}
  \label{fig:visibles}
\end{figure}

~\\[-1cm]
{\bfseries Joint distribution} \\[.25cm]
Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM. Then  the probability corresponding to the state of each node taking the value of $1$:
\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\Nsum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)} 
\end{align}

@@

%%-----------------------------------------------
%%
%%Middle left box
%%
\begin{textblock}{14}(.65, 11.33)
{\large \bfseries Degeneracy, instability, and uninterpretability... Oh my!}
~\\[.25cm]
\begin{table}
\centering
\begin{tabular}{| p{.48\linewidth} | p{.5\linewidth} |}
\hline
Characteristic & Detection \\
\hline
\hline
{\bfseries Near-degeneracy.} Occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space by the model \cite{handcock2003assessing}. & If the mean parametrization on the model parameters, $\mu(\boldsymbol \theta)$, is close to the boundary of the convex hull of the set of statistics in the neg-potential function $Q(\boldsymbol x)$. \\
\hline
{\bfseries Instability.} Small changes in natural parameters result in large changes of the pmf, excessive sensitivity \cite{schweinberger2011instability}. & If for any $C > 0$ there exists $N_C > 0$ such that 
$\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$. Where $Q(\cdot)$ is the neg-potential function of the model. \\
\hline
{\bfseries Uninterpretability.} Due to the existence of dependence, marginal mean-structure no longer maintained \cite{kaiser2007statistical}. & If the magnitude of the difference between model expectations and expectations under independence, $\left|\text{E}(\boldsymbol X | \boldsymbol \theta) - \text{E}(\boldsymbol X | \emptyset ) \right|$, is large.\\
\hline
\end{tabular}
\label{tab:degen}
\caption{Table of ``bad" model characteristics.}
\end{table}
\end{textblock}

%%-----------------------------------------------
%%
%%Bottom left box
%%
\begin{textblock}{14}(.65, 22.94)
{\large \bfseries Variable encoding}

@@

%%-----------------------------------------------
%%
%%Top right box
%%
\begin{textblock}{14}(32.4, 5.71)
{\large \bfseries Managable examples} \\[.25cm]
\end{textblock}

@@

\begin{textblock}{14.54}(31.86, 24.6)
These models very easily are degenerate, unstable, and uninterpretable. Meaning, the space of possible fitted parameter values that leads to a proper model is highly restricted.

@@

%%-----------------------------------------------
%%
%%Bottom middle box
%%
\begin{textblock}{14.5}(16.17, 24.84)
{\large \bfseries References}
\printbibliography[heading=none]

>>>>

%%tikz setup
\usepackage{tikz, subfig, amsthm}
\usepackage{tikz-3dplot}
\usetikzlibrary{arrows, shapes, positioning, backgrounds, decorations.pathreplacing}

%%other options
\definecolor{isublue}{RGB}{58,117,196}
\definecolor{isured}{RGB}{206,17,38}
\definecolor{isugreen}{RGB}{7,109,84}
\counterwithout{figure}{chapter}
\counterwithout{table}{chapter}
\usepackage{amsmath}
\setlength{\parindent}{0cm}
\def\vdotswide{\vbox{\baselineskip=20pt \lineskiplimit=0pt \kern6pt \hbox{.}\hbox{.}\hbox{.}}}

@@

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\input{images/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), which consists of two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer. Hidden nodes indicated by white circles and the visible nodes indicated by blue circles \cite{friedman2001elements}.}
  \label{fig:rbm}
\end{figure}

\vspace{1cm}
\begin{figure}[ht]
  \centering
  \input{images/visibles.tikz}
  \caption{Image classification using a RBM. On the left, each image pixel comprises a node in the visible layer, $\mathcal{V}$. On the right, the output of the RBM is used to create features which are then passed to a supervised learning algorithm.} 
  \label{fig:visibles}
\end{figure}

~\\[-1cm]
{\bfseries Joint distribution} \\[.25cm]
Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM. Then the probability each node taking the the value corresponding to $\boldsymbol x$ is:
\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\Nsum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)} 
\end{align}

@@

%%-----------------------------------------------
%%
%%Middle left box
%%
\begin{textblock}{14}(.65, 11.33)
{\large \bfseries Degeneracy, instability, and uninterpretability... Oh my!}\\[-1cm]

The highly flexible nature of the RBM ($H + V + HV$ parameters) makes the following characteristics of model impropriety of particular concern. \\[-.5cm]

\begin{table}
\centering
\begin{tabular}{| p{.48\linewidth} | p{.5\linewidth} |}
\hline
Characteristic & Detection \\
\hline
\hline
{\bfseries Near-degeneracy.} Occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space by the model \cite{handcock2003assessing}. & If the mean parametrization on the model parameters, $\mu(\boldsymbol \theta)$, is close to the boundary of the convex hull of the set of statistics in the neg-potential function $Q(\boldsymbol x)$. \\
\hline
{\bfseries Instability.} Small changes in natural parameters result in large changes of the pmf, excessive sensitivity \cite{schweinberger2011instability}. & If for any $C > 0$ there exists $N_C > 0$ such that 
$\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$. Where $Q(\cdot)$ is the neg-potential function of the model. \\
\hline
{\bfseries Uninterpretability.} Due to the existence of dependence, marginal mean-structure no longer maintained \cite{kaiser2007statistical}. & If the magnitude of the difference between model expectations and expectations under independence, $\left|\text{E}(\boldsymbol X | \boldsymbol \theta) - \text{E}(\boldsymbol X | \emptyset ) \right|$, is large.\\
\hline
\end{tabular}
\label{tab:degen}
\caption{Table of ``improper model'' characteristics.}
\end{table}
\end{textblock}

%%-----------------------------------------------
%%
%%Bottom left box
%%
\begin{textblock}{14}(.65, 21.5)
{\large \bfseries Avoiding degeneracy} \\[-.75cm]

For the $\{-1, 1 \}$ encoding of $\mathcal{V}$ and $\mathcal{H}$, the origin is the center of the parameter space. In particular, at $\boldsymbol \theta = \boldsymbol 0$, the RBM is equivalent to elements of $\boldsymbol X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ r.v.s. $\Rightarrow$ No \emph{near-degeneracy}, \emph{instability}, or \emph{uninterpretability}!
\end{textblock}
\begin{textblock}{14.5}(.65, 23.9)
\begin{minipage}{.42\textwidth}
\begin{figure}[ht]
  \begin{minipage}{0.49\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{-60}
    \input{images/toyhull_top.tikz}

    \draw(0,1,0) -- (1,0,0) -- (0,0,0) -- (0,1,0); 
     \draw(1,1,1) -- (1,0,0) -- (0,0,0) -- (1,1,1); 
     \draw(1,1,1) -- (0,1,0) -- (0,0,0) -- (1,1,1); 
     \draw(1,1,1) -- (0,1,0) -- (1,0,0) -- (1,1,1); 
    \end{tikzpicture}
  }
  \end{minipage}
  \begin{minipage}{0.49\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{-60}
    \input{images/toyhull_top_2.tikz}

    \draw(-1,1,-1) -- (1,-1,-1) -- (-1,-1,-1) -- (-1,1,-1); 
     \draw(1,1,1) -- (1,-1,-1) -- (-1,-1,-1) -- (1,1,1); 
     \draw(1,1,1) -- (-1,1,-1) -- (-1,-1,-1) -- (1,1,1); 
     \draw(1,1,1) -- (-1,1,-1) -- (1,-1,-1) -- (1,1,1); 
    \end{tikzpicture}
  }
  \end{minipage}
  \caption{The convex hulls of the statistic space in three dimensions for a toy RBM with $|\mathcal{V}| = |\mathcal{H}| = 1$ for $\{0,1\}$-encoding (left) and $\{-1,1\}$-encoding (right) enclosed by an unrestricted hull of 3-space.}
\label{fig:toyhull}
\end{figure}
\end{minipage}
\begin{minipage}{.02\textwidth}
\hfill
\end{minipage}
\begin{minipage}{.55\textwidth}
\begin{figure}
\centering
\includegraphics[scale=1]{images/frac_volume.pdf}
\label{fig:frac_volume}
\caption{Volume relationship for the convex hulls of statistics in $Q(\cdot)$ vs. unrestricted space.}
\end{figure}
\end{minipage}

@@

%%-----------------------------------------------
%%
%%Top right box
%%
\begin{textblock}{14}(32.4, 5.71)
{\large \bfseries Manageable (a.k.a. small) examples} \\[.25cm]

As the magnitude of $\boldsymbol \theta$ grows, so does the occurence of near-degeneracy, instability, and uninterpretability for RBMs of varying sizes.

@@

\begin{textblock}{14.54}(31.86, 24.8)
These models very easily are degenerate, unstable, and uninterpretable. Meaning, the space of possible fitted parameter values that leads to a proper model is highly restricted.

@@

%%-----------------------------------------------
%%
%%Bottom middle box
%%
\begin{textblock}{14.5}(16.17, 25.45)
{\large \bfseries References}
\printbibliography[heading=none]

### 2016-02-20 writing/degeneracy/draft.Rmd
left side complete

---
title: "Degeneracy in a restricted Boltzmann machine"
author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
date: "January 8, 2016"
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: ../references/refs.bib
---

>>>>

---
title: "Degeneracy in a restricted Boltzmann machine"
author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
date: "January 8, 2016"
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    number_sections: yes
bibliography: ../references/refs.bib
---
